---
title: "Public and economic effects of storms and other severe weather events in USA"
author: "Aanjaneya Mishra"
date: "16/02/2015"
output: html_document
keep_md: true
---

```{r echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Synopsis

In this report we explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. While exploring the database we will be answering the following: 

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing

From the URL [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) we obtained U.S. National Oceanic and Atmospheric Administration's (NOAA) data on major storms and weather events in the United States from the year 1950 to November 2011. 

### Downloading the data

```{r}
if(!file.exists("repdata_data_StormData.csv.bz2")) {
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileURL, destfile="repdata_data_StormData.csv.bz2", method = "curl")
}
```

### Reading The data

```{r}
stormdata <- read.csv(bzfile("repdata_data_StormData.csv.bz2"))
```
### Data summary

```{r}
dim(stormdata)
str(stormdata)
```

### Processing the data for the purpose of getting results

#### Effect of storms and severe weather on health of population accross USA

To be able to determine this aspect in the results section we will use the FATALITIES and INJURIES columns of the storm data as these columns are relavant for determination of effect on public health. 

Aggregate and order fatalities by event type

```{r}
fatalities <- aggregate(FATALITIES ~ EVTYPE, stormdata, sum)
colnames(fatalities) <- c("Event", "Fatalaties")
fatalities <- fatalities[order(fatalities$Event), ]
```

Aggregate and order injuries by event type

```{r}
injuries <- aggregate(INJURIES ~ EVTYPE, stormdata, sum)
colnames(injuries) <- c("Event", "Injuries")
injuries <- injuries[order(injuries$Event), ]
```

Combine fatalities and injuries by event type.

```{r}
healthimpact <- cbind(fatalities, injuries$Injuries)
colnames(healthimpact) <- c("Event", "Fatalaties", "Injuries")
```

Clean and narrow down healthimpact table by removing values where fatalities and injuries are 0. We do this since events with 0 occurences are not useful for determination of meaningful results.

```{r}
healthimpact <-healthimpact[healthimpact$Fatalaties > 0 | healthimpact$Injuries > 0, ]
# rownames(healthimpact) <- NULL
head(healthimpact)
```

We order the healthimpact table to put the events in decreasing order of effect on population health. In other words decreasing ordering will put most harmful events at the top of the table while those least harmful at the bottom.

```{r}
healthimpact <- healthimpact[order(healthimpact$Fatalaties, healthimpact$Injuries, decreasing = TRUE), ]
head(healthimpact)
```

#### Economic effects of storms and severe weather accross USA

To be able to determine this aspect in the results section we will use the PROPDMG, PROPDMGEXP, CROPDMG and CROPDMGEXP columns of the storm data as these columns are relavant for determination economic consequences. The columns PROPDMG and CROPDMG contain units of damage whereas the PROPDMGEXP and CROPDMGEXP contain exponentials(10^x) like million or billion to be applied to PROPDMG and CROPDMG. Therefore, in order to get the true value of property and crop damage we need to interpret, recode the PROPDMGEXP and CROPDMGEXP variables and then merge them with PROPDMG and CROPDMG.

Interpret and recode the PROPDMGEXP and CROPDMGEXP values to numeric versions so that we can apply arithmetic calculations to it.

```{r}
library(car)
stormdata$PROPDMGEXP <- factor(tolower(stormdata$PROPDMGEXP))
stormdata$PROPDMGEXP <- as.numeric(recode(as.character(stormdata$PROPDMGEXP), "'0'=1;'1'=10;'2'=10^2;'3'=10^3;'4'=10^4;'5'=10^5;'6'=10^6;'7'=10^7;'8'=10^8;'b'=10^9;'h'=10^2;'k'=10^3;'m'=10^6;'-'=0;'?'=0;'+'=0"))

stormdata$CROPDMGEXP <- factor(tolower(stormdata$CROPDMGEXP))
stormdata$CROPDMGEXP <- as.numeric(recode(as.character(stormdata$CROPDMGEXP), "'0'=1;'1'=10;'2'=10^2;'3'=10^3;'4'=10^4;'5'=10^5;'6'=10^6;'7'=10^7;'8'=10^8;'b'=10^9;'h'=10^2;'k'=10^3;'m'=10^6;'-'=0;'?'=0;'+'=0"))
```

Apply the recoded numeric values of PROPDMGEXP and CROPDMGEXP to PROPDMG and CROPDMG to calculate the actual crop and property damages. It will be done by multiplying PROPDMGEXP with PROPDMG and CROPDMGEXP with CROPDMG.

```{r}
stormdata$PROPDMGVALUE <- stormdata$PROPDMG * stormdata$PROPDMGEXP
stormdata$CROPDMGVALUE <- stormdata$CROPDMG * stormdata$CROPDMGEXP
```


Aggregate and order property damage by event type

```{r}
propdamage <- aggregate(PROPDMGVALUE ~ EVTYPE, stormdata, sum)
colnames(propdamage) <- c("Event", "PropertyDamage")
propdamage <- propdamage[order(propdamage$Event), ]
```

Aggregate and order crop damage by event type

```{r}
cropdamage <- aggregate(CROPDMGVALUE ~ EVTYPE, stormdata, sum)
colnames(cropdamage) <- c("Event", "CropDamage")
cropdamage <- cropdamage[order(cropdamage$Event), ]
```

Clean and narrow down propdamage and cropdamage tables by removing values where propertydamage and cropdamage are 0. We do this since events with 0 occurences are not useful for determination of meaningful results.

```{r}
propimpact <- propdamage[propdamage$PropertyDamage > 0, ]
head(propimpact)
cropimpact <- cropdamage[cropdamage$CropDamage > 0, ]
head(cropimpact)
```

We order the propimpact and cropimpact tables to put the events in decreasing order of economic effect. In other words decreasing ordering will put most economicall harmful events at the top of the table while those least harmful at the bottom.

```{r}
propimpact <- propimpact[order(propimpact$PropertyDamage, decreasing = TRUE), ]
head(propimpact)
cropimpact <- cropimpact[order(cropimpact$CropDamage, decreasing = TRUE), ]
head(cropimpact)
```

## Results

### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

To answer this question we will draw a stacked barplot to see the fatalities and injuries related to different most devastating weather events. We will also see the numbers (numeric view )of faltalities and injuries for top 6 events.

Numeric view of faltalities and injuries for top 6 events. We can see that Tornados are most harmful with respect to population health. Tornados are followed by Excessive Heat, Flash Flood, Heat. Lightning etc.

```{r}
head(healthimpact)
```

To draw the barplot we need to convert heathimpact table calculated above into matrix and transpose it.

```{r}
plotdatahealth <- as.matrix(healthimpact[, c("Fatalaties", "Injuries")])
rownames(plotdatahealth) <- healthimpact$Event
tplotdatahealth <- t(plotdatahealth)
```

Now the bar plot

```{r}
barplot(tplotdatahealth, names.arg = healthimpact$Event, legend = c( "Fatalities", "Injuries"), 
    main = "Total fatalities and injuries by events", col = c("red", "darkblue"), xlim = c(0,10), ylim = c(0,10000))
```

